Definition of Archipelagos

### 2025-01-04 application_examples/abc_inferences/archipelago-definition.Rmd
Adding result plots for CFTP convergence test

```{r}
# If missing, install required packages

@@

library(rtracklayer)
```

@@

## Definition of promoter regions

@@

```{r}
# Extract TSS (assuming + strand)
promoters <- data.frame(
  chr = seqnames(gtf_transcripts),
  tss = ifelse(strand(gtf_transcripts) == "+", start(gtf_transcripts), end(gtf_transcripts)),
  strand = strand(gtf_transcripts)
)
```

>>>>

```{r, warning=FALSE, message=FALSE}
# If missing, install required packages

@@

if (!requireNamespace("GenomicRanges", quietly = TRUE)) {
  BiocManager::install("GenomicRanges")
}

library(GenomicRanges)
library(rtracklayer)
```

@@

## Step 1: Definition of promoter regions

@@

```{r}
# Extract TSS
promoters <- data.frame(
  chr = seqnames(gtf_transcripts),
  tss = ifelse(strand(gtf_transcripts) == "+", start(gtf_transcripts), end(gtf_transcripts)),
  strand = strand(gtf_transcripts)
)
head(promoters)
```

The promoters data frame contains:

- `chr`: Chromosome names

- `tss`: Transcription start sites

- `strand`: Strand orientation of the transcripts

## Step 2: Filtering of CGI by promoter overlap

CGI annotation file (`.txt` file) obtained with makeCGI v1.3.4 [@hiddenMarkov]

```{r}
# Load the CGI annotation file
cgi_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/CGI-Mzebra.txt" 
cgif <- read.table(cgi_file, header = TRUE, sep = "\t")
head(cgif)
```

Convert the promoter and CGI data to `GRanges` objects for genomic interval operations:

```{r}
# Create GRanges for promoters
promoter_gr <- GRanges(
  seqnames = promoters$chr,
  ranges = IRanges(start = promoters$tss, end = promoters$tss),
  strand = promoters$strand
)

# Create GRanges for CGI
cgi_gr <- GRanges(
  seqnames = cgif$chr,
  ranges = IRanges(start = cgif$start, end = cgif$end)
)
```

Use `findOverlaps` to find promoters within or near CpG islands:

```{r}
# Find overlaps
overlaps <- findOverlaps(promoter_gr, cgi_gr)

# Extract overlapping promoter indices
promoters_with_cgi <- promoters[queryHits(overlaps), ]
cgi_overlaps <- cgif[subjectHits(overlaps), ]

# Combine data for output
promoter_cgi_overlap <- cbind(promoters_with_cgi, cgi_overlaps)
head(promoter_cgi_overlap)
```

The `promoter_cgi_overlap` dataset combines information about promoters and CpG islands (CGIs) where they overlap. It contains the following columns:

- `chr`: The chromosome where the promoter is located.

- `tss`: The transcription start site (TSS) position of the promoter.

- `strand`: The strand of the promoter (+ or -).

- `chr`: The chromosome where the CGI is located (same as the promoter in the overlap context).

- `start`: The start position of the overlapping CGI.

- `end`: The end position of the overlapping CGI.

- `length`: The length of the CGI in base pairs.

- `CpGcount`: The number of CpG dinucleotides within the CGI.

- `GCcount`: The number of G and C nucleotides within the CGI.

- `pctGC`: The percentage of G and C nucleotides within the CGI.

- `obsExp`: The observed-to-expected ratio of CpG dinucleotides in the CGI.

Each row corresponds to an overlap between a promoter and a CGI. For example:

- The first row shows that a promoter at TSS 755,445 on the negative strand overlaps with a CGI on chromosome NC_036780.1 spanning positions 755,397 to 756,689. This CGI has a length of 1,293 bp, contains 123 CpG dinucleotides, and has a GC content of 63.19% with an observed-to-expected CpG ratio of 0.954.

- The overlap is detailed for both the promoter and the CGI, including their respective properties.

```{r}
#write.table(promoter_cgi_overlap, file = "promoter_cgi_overlap.txt", sep = "\t", row.names = FALSE)
promoter_cgi_overlap <- read.table("/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/promoter_cgi_overlap.txt")
```

### Check TSS overlap density

```{r}
plot(
  promoter_cgi_overlap$length, 
  promoter_cgi_overlap$pctGC, 
  xlab = "CGI Length (bp)", 
  ylab = "GC Content (%)", 
  main = "Relationship Between CGI Length and GC Content",
  pch = 16,  # Solid circle points
  col = "blue"  # Point color
)

```

```{r}

```
